mathml: Translate R expressions to MathML and LaTeX/MathJax

### 2022-12-11 Draft2.Rmd
hooks

---
title: "mathml: Translate R expressions to MathML and LaTeX/MathJax"
date: "2022-11-09"
abstract: >
  This R package translates R objects to suitable entities in MathML or LaTeX,
  thereby enabling pretty mathematical representation of R functions or
  algorithms in data analysis, scientific reports and interactive web content.
  R code text descriptions and mathematical content already coexist,
  side-by-side, in the RMarkdown document rendering language. The present
  package allows to use the same R\ objects for both data analysis and
  typesetting in documents or web content. By tightening the link between the
  statistical analysis and its verbal or symbolic description, this new feature
  is a futher step towards reproducible science. The package can be extended by
  user-defined hooks that allow to replace variable names and new R functions by
  a suitable symbolic representation. We provide two working examples that
  illustrate the use of the new package in a scientific article and in an
  interactive web page.
draft: true
author:  
  # see ?rjournal_article for more information
  - name: Matthias Gondan
    affiliation: University of Innsbruck
    address:
    - Department of Psychology
    - Innsbruck, Austria
    url: https://www.uibk.ac.at/psychologie/mitarbeiter/gondan-rochon/index.html.en
    orcid: 0000-0001-9974-0057
    email:  Matthias.Gondan-Rochon@uibk.ac.at
  - name: Irene Alfarone
    url: https://www.uibk.ac.at/psychologie/mitarbeiter/alfarone/index.html.en
    email: Irene.Alfarone@uibk.ac.at
    orcid: 0000-0002-8409-8900
    affiliation: University of Innsbruck
    address:
    - Department of Psychology
    - Innsbruck, Austria
type: package
output: 
  rjtools::rjournal_web_article:
    self_contained: yes
    toc: no
bibliography: RJreferences.bib

---

>>>>

---
title: "mathml: Translate R expressions to MathML and LaTeX/MathJax"
date: "2022-11-09"
abstract: >
  This R package translates R objects to suitable entities in MathML or LaTeX,
  thereby enabling pretty mathematical representation of R functions or
  algorithms in data analysis, scientific reports and interactive web content.
  R code text descriptions and mathematical content already coexist,
  side-by-side, in the RMarkdown document rendering language. The present
  package allows to use the same R\ objects for both data analysis and
  typesetting in documents or web content. By tightening the link between the
  statistical analysis and its verbal or symbolic description, this new feature
  is a futher step towards reproducible science. The package can be extended by
  user-defined hooks that allow to replace variable names and new R functions by
  a suitable symbolic representation. We provide two working examples that
  illustrate the use of the new package in a scientific article and in an
  interactive web page.
draft: true
author:  
  # see ?rjournal_article for more information
  - name: Matthias Gondan
    affiliation: University of Innsbruck
    address:
    - Department of Psychology
    - Innsbruck, Austria
    url: https://www.uibk.ac.at/psychologie/mitarbeiter/gondan-rochon/index.html.en
    orcid: 0000-0001-9974-0057
    email:  Matthias.Gondan-Rochon@uibk.ac.at
  - name: Irene Alfarone
    url: https://www.uibk.ac.at/psychologie/mitarbeiter/alfarone/index.html.en
    email: Irene.Alfarone@uibk.ac.at
    orcid: 0000-0002-8409-8900
    affiliation: University of Innsbruck
    address:
    - Department of Psychology
    - Innsbruck, Austria
type: package
output: 
  rjtools::rjournal_web_article:
    self_contained: yes
    toc: no
bibliography: RJreferences.bib
---

### 2022-12-11 Draft3.Rmd
hooks

Note that, on the Prolog end, the handler for `integrate/3` is rather rigid; it
accepts only these three arguments in that particular order, and without names,
that is, `integrate(sin, lower=0L, upper=2L*pi)`, would not print the desired
result.

@@

# Package capabilities

## Hooks
Before start listing all the possible operations with \pkg{mathml},  we mention here that \pkg{mathml} also allows the user to create hooks for specific functions. As an example:

```{r, echo=TRUE}
# Improve display
hook(quote(mu_A), quote(mu["A"]))
hook(quote(mu_B), quote(mu["B"]))
hook(quote(sigma_A), quote(sigma["A"]))
hook(quote(sigma_B), quote(sigma["B"]))
hook(quote(mu_M), quote(mu["M"]))
hook(quote(zn), quote(z[tau]))
hook(quote(M), quote(overline(X)))
```

The hooks reported here will be the one used in Section _ref_ to display the equations in the Case Study. On the right hand side of the parenthesis is reported the way that the variable, for instance

>>>>

## Names and order of arguments

One of R's great features is the possibility to refer to function arguments by
by their names, not only by their position in the list of arguments.
Unfortunately, the prolog-powered handler for `integrate/3` is rather rigid; it
accepts only three arguments in that particular order and without names, that
is, `integrate(lower=0L, upper=2L*pi, sin)`, would not print the desired result.

@@

## Customizing the package

In typical R\ functions, variable names are typically longer than just single
letters, which may yield unsatisfactory results in the mathematical output.

```{r, results='asis'}
term <- quote(pbinom(successes, Ntotal, prob))
mathout(term)
```

To improve the situation, \pkg{mathml} provides a simple hook into the prolog
predicate `math/2` described in the previous section.

```{r, results='asis'}
hook(quote(successes), quote(k))
hook(quote(Ntotal), quote(N))
hook(quote(prob), quote(pi))
mathout(term)
```

More complex hooks are possible. In the example below, `dnorm(X, M, S)` is
translated to the form `varphi(X; M, S)`.

```{r, results='asis'}
hook(call("dnorm", expression(X), expression(M), expression(S)), 
     call("fn", quote(varphi),
       call(";", list(expression(X)), list(expression(M), expression(S)))))
term <- quote(dnorm(x, mu, sigma^2L))
mathout(term)
```

R's expressions are a bit misused to represent variables in the underlying
prolog engine.
